---
title: "STAT 547 Project Code"
output: html_document
date: "2026-02-20"
---
```{r}
# adjust to your setting
.libPaths("~/Rlib/")
setwd("~/Desktop/Msc Statistics/2025 Winter Term 2/STAT 547/547 Project")

library(outbreaks)
library(ggplot2)
theme_set(theme_bw())
library(rjson)
library(cmdstanr)

library(bayesplot)
library(posterior)
library(loo)
library(tidyverse)

source("tools_is.r")
library(deSolve)
library(ggplot2)
library(tidyr)
cmdstanr::set_cmdstan_path("/Users/jessieliu/Desktop/Msc Statistics/2025 Winter Term 2/STAT 547/Course code/Torsten/Torsten/cmdstan")
cmdstanr::cmdstan_path()
cmdstanr::cmdstan_version()
```

# Data simulation

## Individual Compartment Model Data Simulation
```{r}
### ONE COMPARTMENT MODEL
one_comp_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    du_gut  <- -ka * u_gut
    du_cent <- ka * u_gut - (CL/Vcent) * u_cent 
    return(list(c(du_gut, du_cent)))
  })
}


### TWO COMPARTMENT MODEL
two_comp_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    du_gut  <- -ka * u_gut
    du_cent <- ka * u_gut - (CL/Vcent + Q/Vcent) * u_cent + (Q/Vperi) * u_peri
    du_peri <- (Q/Vcent) * u_cent - (Q/Vperi) * u_peri
    return(list(c(du_gut, du_cent, du_peri)))
  })
}



### PARAMETERS
par_comp_model <- function(num_comp, seed){
  set.seed(seed)
  if(num_comp == 1){
    comp_params <- data.frame(
    CL     = rlnorm(1, log(10), 0.25), # clearance rate at central compartment
    Vcent  = rlnorm(1, log(35), 0.25), # volume of central compartment
    ka     = rlnorm(1, log(2.5), 1), # absorption rate 
    sigma_res <- abs(rnorm(1, 0, 1))) # standard deviation of observed data: concentration at central compartment
  
  }else{
    comp_params <- data.frame(
    CL     = rlnorm(1, log(10), 0.25),
    Q      = rlnorm(1, log(15), 0.5), # interchange rate between compartments
    Vcent  = rlnorm(1, log(35), 0.25),
    Vperi  = rlnorm(1, log(105), 0.5),
    ka     = rlnorm(1, log(2.5), 1),
    sigma_res <- abs(rnorm(1, 0, 1)))
  }

  return(comp_params)
}



### Observed data: Event Schedule + Dose Schedule
simulated_data <- function(num_comp, parameter, dose = 1200, seed){
  ## Time points
  intensive_pts <- c(0.083, 0.167, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8) # blood draw after dose
  t_dose1 <- intensive_pts
  t_dose2 <- 12 + intensive_pts
  t_dose14 <- 156 + intensive_pts
  t_trough <- dose_times <- seq(0, 156, by = 12) # blood draw before dose (assume drug treatment and blood draw happens at the same time)

  t_washout <- 156 + c(12, 18, 24) # blood draw after the last dose
  all_obs_times <- sort(unique(c(t_dose1, t_dose2, t_dose14, t_trough, t_washout)))
  
  
  ## Drug Concentration at central compartment
   events_df <- data.frame(
    var    = "u_gut",       
    time   = dose_times,    
    value  = dose,          
    method = "add"          
  )
  
  if (num_comp == 1) {y <- c(u_gut = 0, u_cent = 0)
  } else {y <- c(u_gut = 0, u_cent = 0, u_peri = 0)
  
} # initialize the drug mass at each compartment is 0
  model <- if (num_comp == 1) one_comp_model else two_comp_model
  out <- as.data.frame(ode(y, 
                           times = all_obs_times, 
                           func = model, 
                           parms = parameter,
                           events = list(data = events_df))) # multiple dose
  
  Vc <- parameter$Vcent
  sig <- parameter$sigma_res

  c_true <- out$u_cent / Vc
  c_true <- pmax(c_true, 1e-12)                 # avoid 0
  cObs <- c_true * exp(rnorm(nrow(out), 0, sig)) # lognormal noise
  
  ## Combine drug treatment with blood draw
  dose_row <- data.frame(
     time = 0, addl = 13, ii = 12, amt = dose, 
    cmt = 1, evid = 1, rate = 0, ss = 0, cObs = NA
  )
  obs_rows <- data.frame(
    time   = out$time,
    addl   = 0,
    ii     = 0,
    amt    = 0,
    cmt    = 2,
    evid   = 0,
    rate   = 0,
    ss     = 0,
    cObs   = cObs
  )
  
  dat <- rbind(dose_row, obs_rows)
  dat <- dat[order(dat$time, -dat$evid), ] 
  return(dat)
}




### CONVERT DATA FOR GAUSSIAN PROCESS

to_gp_data <- function(dat, dose = 1200) {
  dat <- dat[order(dat$time, -dat$evid), ]
  rownames(dat) <- NULL

  iObs <- which(dat$evid == 0)

  obs <- dat[iObs, ] %>% arrange(time)

  list(
    N = nrow(obs),
    t_obs = as.numeric(obs$time),
    y_obs = as.numeric(obs$cObs),
    dose = dose
  )
}



### CONVERT DATA FOR PK MODEL

to_torsten_data <- function(dat) {
  # 保证同一时刻 dose 在 obs 前
  dat <- dat[order(dat$time, -dat$evid), ]
  rownames(dat) <- NULL
  
  iObs <- which(dat$evid == 0)  # 观测行在 1:nt 里的索引
  
  list(
    nt   = nrow(dat),
    nObs = length(iObs),
    iObs = as.integer(iObs),
    
    cmt  = as.integer(dat$cmt),
    evid = as.integer(dat$evid),
    addl = as.integer(dat$addl),
    ss   = as.integer(dat$ss),
    
    amt  = as.numeric(dat$amt),
    time = as.numeric(dat$time),
    rate = as.numeric(dat$rate),
    ii   = as.numeric(dat$ii),
    
    cObs = as.vector(pmax(dat$cObs[iObs], 1e-12))  # 只给观测值，且>0
  )
}

```


Data Simulation
```{r}
num_replicates <- 5
seeds <- sample(1:10000, num_replicates, replace = FALSE)
datasets_1c <- lapply(seeds, function(s) {
  p <- par_comp_model(num_comp = 1, seed = s)  # parameters for individua
  dat <- simulated_data(
    num_comp = 1,
    parameter = p,
    dose = 1200,
    seed = s
  )
  dat$seed <- s   # label/index for each dataset
  dat
})



datasets_2c <- lapply(seeds, function(s) {
  p <- par_comp_model(num_comp = 2, seed = s)
  dat <- simulated_data(
    num_comp = 2,
    parameter = p,
    dose = 1200,
    seed = s
  )
  dat$seed <- s
  dat
})


gp_data_list_1c <- lapply(datasets_1c, to_gp_data, dose = 1200)
gp_data_list_2c <- lapply(datasets_2c, to_gp_data, dose = 1200)


pk_data_list_1c <- lapply(datasets_1c, to_torsten_data)
pk_data_list_2c <- lapply(datasets_2c, to_torsten_data)
```




# Run and fit model

## PK model
```{r}
mod_pk <- cmdstan_model("two_compartment.stan")

init_2c <- function() list(
  CL = 10, Q = 15, V1 = 35, V2 = 105, ka = 2.5, sigma = 0.3
)

# Mis-specified case for pk2 model: Using pk2 model to fit pk1 data
fit_pk_1c <- lapply(pk_data_list_1c, function(pk_data) {
  mod_pk$sample(
    data = pk_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = init_2c,
    refresh = 50,
    show_messages = FALSE,
    show_exceptions = FALSE
  )
})

fit_pk_2c <- lapply(pk_data_list_1c, function(pk_data) {
  mod_pk$sample(
    data = pk_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = init_2c,
    refresh = 0,
    show_messages = FALSE,
    show_exceptions = FALSE
  )
})


```

### Inference (convergence, eﬀective sample size) for PK model
```{r}
pars_pk <- c("CL","Q","V1","V2","ka","sigma")

# Diagnostic for one-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_pk_1c), function(i) {
  s <- fit_pk_1c[[i]]$summary(variables = pars_pk)
  s$dataset_id <- i
  s
}))
subset(summ_table, rhat > 1.01 | ess_bulk < 200 | ess_tail < 200)

# Diagnostic for two-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_pk_2c), function(i) {
  s <- fit_pk_2c[[i]]$summary(variables = pars_pk)
  s$dataset_id <- i
  s
}))
subset(summ_table, rhat > 1.01 | ess_bulk < 200 | ess_tail < 200)
```

### Posterior predictive check for PK model
```{r}
compute_loo <- function(fit, log_lik_name = "log_lik") {
  # compute log_likihood draws
  ll_array <- fit$draws(variables = log_lik_name, format = "draws_array")
  ll_mat <- posterior::as_draws_matrix(ll_array)
  loo::loo(ll_mat)
}

loo_pk_1c <- lapply(fit_pk_1c, compute_loo, log_lik_name = "log_lik")
loo_pk_2c <- lapply(fit_pk_2c, compute_loo, log_lik_name = "log_lik")
```




## Gaussian Process
```{r}
mod_gp <- cmdstan_model("gp_model.stan")
make_init_gp <- function(gp_data) {
  function() list(
    a     = mean(log(gp_data$y_obs)),
    rho   = 1, #smoothness
    alpha = 1, #fluctuation 
    sigma = 0.3, #latent standard normal
    z     = rep(0, gp_data$N)
  )
}


fit_gp_1c <- lapply(gp_data_list_1c, function(gp_data) {
  mod_gp$sample(
    data = gp_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = make_init_gp(gp_data),
    refresh = 0,
    show_messages = FALSE,
  show_exceptions = FALSE
  )
})

fit_gp_2c <- lapply(gp_data_list_2c, function(gp_data) {
  mod_gp$sample(
    data = gp_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = make_init_gp(gp_data),
    refresh = 0,
    show_messages = FALSE,
  show_exceptions = FALSE
  )
})


```



### Inference (convergence, eﬀective sample size)
```{r}
pars_gp <- c("a", "rho", "alpha", "sigma")

# Diagnostic for one-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_gp_1c), function(i) {
  s <- fit_gp_1c[[i]]$summary(variables = pars_gp)
  s$dataset_id <- i
  s
}))
# Filter: Rhat > 1.01
subset(summ_table, rhat > 1.01)
# Filter: ESS < 200
subset(summ_table, ess_bulk < 200 | ess_tail < 200)


# Diagnostic for two-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_gp_2c), function(i) {
  s <- fit_gp_2c[[i]]$summary(variables = pars_gp)
  s$dataset_id <- i
  s
}))
# Filter: Rhat > 1.01
subset(summ_table, rhat > 1.01)
# Filter: ESS < 200
subset(summ_table, ess_bulk < 200 | ess_tail < 200)


```


### Posterior predictive check (LOO Diagnostics)

```{r}
loo_gp_1c <- lapply(fit_gp_1c, compute_loo, log_lik_name = "log_lik")
loo_gp_2c <- lapply(fit_gp_2c, compute_loo, log_lik_name = "log_lik")

```
There is an outlier in the simulated data has very large shape parameter $k$. Need to pinpoint the point and do exact LOO refit.

“最差的总是第 1 个数据点”其实非常合理：你 GP 用的观测数据里，第 1 个观测点几乎一定是 time = 0 那个 trough 点，而这个点在你现在的模拟机制下会天然变成 高杠杆 / 高敏感点，PSIS-LOO 就很容易把它标成最麻烦的点（Pareto k 最大）。
```{r}
loo::pareto_k_table(loo_gp_1c[[1]])
loo::pareto_k_table(loo_gp_2c[[1]])


k <- loo_gp_1c[[1]]$diagnostics$pareto_k
i_bad <- which(k > 1)
i_bad

```
```{r}

# Extract posterior predictive checks
# Gaussian Process
yrep_draws <- fit_gp_1c[[1]]$draws(variables = "y_rep")
yrep_mat <- as_draws_matrix(yrep_draws)
idx <- as.integer(gsub("y_rep\\[|\\]", "", colnames(yrep_mat)))
yrep_mat <- yrep_mat[, order(idx), drop = FALSE]


bayesplot::ppc_ribbon(
  y    = datasets_1c[[1]]$cObs[-1],
  yrep = yrep_mat,
  x    = datasets_1c[[1]]$time[-1]
) +
  theme_bw() +
  labs(
    title = "GP Model PPC",
    y = "Concentration",
    x = "Time (h)"
  )


# PK model
yrep_draws <- fit_pk_1c[[1]]$draws(variables = "y_rep")
yrep_mat <- as_draws_matrix(yrep_draws)
idx <- as.integer(gsub("y_rep\\[|\\]", "", colnames(yrep_mat)))
yrep_mat <- yrep_mat[, order(idx), drop = FALSE]


bayesplot::ppc_ribbon(
  y    = datasets_1c[[1]]$cObs[-1],
  yrep = yrep_mat,
  x    = datasets_1c[[1]]$time[-1]
) +
  theme_bw() +
  labs(
    title = "PK Model PPC",
    y = "Concentration",
    x = "Time (h)"
  )

  
```





```{r}
# One compartment data
  # compute PSIS-loo estimate
  
  #1. PSIS-loo estimate of one-compartmental model
  log_lik_1 <- fit_gp_1c[[1]]$draws("log_lik")
  loo1 <- loo(log_lik_1, r_eff = relative_eff(log_lik_1))
  
  #1. PSIS-loo estimate of two-compartmental model
  log_lik_2 <- fit_pk_1c[[1]]$draws("log_lik")
  loo2 <- loo(log_lik_2, r_eff = relative_eff(log_lik_2))
  
  print(loo1)
  print(loo2)
  loo_compare(loo1, loo2)
```








