---
title: "STAT 547 Project Code"
output: html_document
date: "2026-02-20"
---
```{r}
# adjust to your setting
.libPaths("~/Rlib/")
setwd("~/Desktop/Msc Statistics/2025 Winter Term 2/STAT 547/547 Project")

library(outbreaks)
library(ggplot2)
theme_set(theme_bw())
library(rjson)
library(cmdstanr)

library(bayesplot)
library(posterior)
library(loo)
library(tidyverse)

source("tools_is.r")
library(deSolve)
library(ggplot2)
library(tidyr)
cmdstanr::set_cmdstan_path("/Users/jessieliu/Desktop/Msc Statistics/2025 Winter Term 2/STAT 547/Course code/Torsten/Torsten/cmdstan")
cmdstanr::cmdstan_path()
cmdstanr::cmdstan_version()
```

# Data simulation

## Individual Compartment Model Data Simulation
```{r}
### ONE COMPARTMENT MODEL
one_comp_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    du_gut  <- -ka * u_gut
    du_cent <- ka * u_gut - (CL/Vcent) * u_cent 
    return(list(c(du_gut, du_cent)))
  })
}


### TWO COMPARTMENT MODEL
two_comp_model <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    du_gut  <- -ka * u_gut
    du_cent <- ka * u_gut - (CL/Vcent + Q/Vcent) * u_cent + (Q/Vperi) * u_peri
    du_peri <- (Q/Vcent) * u_cent - (Q/Vperi) * u_peri
    return(list(c(du_gut, du_cent, du_peri)))
  })
}



### PARAMETERS
par_comp_model <- function(num_comp, seed){
  set.seed(seed)
  if(num_comp == 1){
    comp_params <- data.frame(
    CL     = rlnorm(1, log(10), 0.25), # clearance rate at central compartment
    Vcent  = rlnorm(1, log(35), 0.25), # volume of central compartment
    ka     = rlnorm(1, log(2.5), 1), # absorption rate 
    sigma_res <- abs(rnorm(1, 0, 1))) # standard deviation of observed data: concentration at central compartment
  
  }else{
    comp_params <- data.frame(
    CL     = rlnorm(1, log(10), 0.25),
    Q      = rlnorm(1, log(15), 0.5), # interchange rate between compartments
    Vcent  = rlnorm(1, log(35), 0.25),
    Vperi  = rlnorm(1, log(105), 0.5),
    ka     = rlnorm(1, log(2.5), 1),
    sigma_res <- abs(rnorm(1, 0, 1)))
  }

  return(comp_params)
}



### Observed data: Event Schedule + Dose Schedule
simulated_data <- function(num_comp, parameter, dose = 1200, seed){
  ## Time points
  intensive_pts <- c(0.083, 0.167, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8) # blood draw after dose
  t_dose1 <- intensive_pts
  t_dose2 <- 12 + intensive_pts
  t_dose14 <- 156 + intensive_pts
  t_trough <- dose_times <- seq(0, 156, by = 12) # blood draw before dose (assume drug treatment and blood draw happens at the same time)

  t_washout <- 156 + c(12, 18, 24) # blood draw after the last dose
  all_obs_times <- sort(unique(c(t_dose1, t_dose2, t_dose14, t_trough, t_washout)))
  
  
  ## Drug Concentration at central compartment
   events_df <- data.frame(
    var    = "u_gut",       
    time   = dose_times,    
    value  = dose,          
    method = "add"          
  )
  
  if (num_comp == 1) {y <- c(u_gut = 0, u_cent = 0)
  } else {y <- c(u_gut = 0, u_cent = 0, u_peri = 0)
  
} # initialize the drug mass at each compartment is 0
  model <- if (num_comp == 1) one_comp_model else two_comp_model
  out <- as.data.frame(ode(y, 
                           times = all_obs_times, 
                           func = model, 
                           parms = parameter,
                           events = list(data = events_df))) # multiple dose
  
  Vc <- parameter$Vcent
  sig <- parameter$sigma_res

  c_true <- out$u_cent / Vc
  c_true <- pmax(c_true, 1e-12)                 # avoid 0
  cObs <- c_true * exp(rnorm(nrow(out), 0, sig)) # lognormal noise
  
  ## Combine drug treatment with blood draw
  dose_row <- data.frame(
     time = 0, addl = 13, ii = 12, amt = dose, 
    cmt = 1, evid = 1, rate = 0, ss = 0, cObs = NA
  )
  obs_rows <- data.frame(
    time   = out$time,
    addl   = 0,
    ii     = 0,
    amt    = 0,
    cmt    = 2,
    evid   = 0,
    rate   = 0,
    ss     = 0,
    cObs   = cObs
  )
  
  dat <- rbind(dose_row, obs_rows)
  dat <- dat[order(dat$time, -dat$evid), ] 
  return(dat)
}




### CONVERT DATA FOR GAUSSIAN PROCESS

to_gp_data <- function(dat, dose = 1200) {
  dat <- dat[order(dat$time, -dat$evid), ]
  rownames(dat) <- NULL

  iObs <- which(dat$evid == 0)

  obs <- dat[iObs, ] %>% arrange(time)

  list(
    N = nrow(obs),
    t_obs = as.numeric(obs$time),
    y_obs = as.numeric(obs$cObs),
    dose = dose
  )
}



### CONVERT DATA FOR PK MODEL

to_torsten_data <- function(dat) {
  # 保证同一时刻 dose 在 obs 前
  dat <- dat[order(dat$time, -dat$evid), ]
  rownames(dat) <- NULL
  
  iObs <- which(dat$evid == 0)  # 观测行在 1:nt 里的索引
  
  list(
    nt   = nrow(dat),
    nObs = length(iObs),
    iObs = as.integer(iObs),
    
    cmt  = as.integer(dat$cmt),
    evid = as.integer(dat$evid),
    addl = as.integer(dat$addl),
    ss   = as.integer(dat$ss),
    
    amt  = as.numeric(dat$amt),
    time = as.numeric(dat$time),
    rate = as.numeric(dat$rate),
    ii   = as.numeric(dat$ii),
    
    cObs = as.vector(pmax(dat$cObs[iObs], 1e-12))  # 只给观测值，且>0
  )
}


```


Data Simulation
```{r}
num_replicates <- 5
seeds <- sample(1:10000, num_replicates, replace = FALSE)
datasets_1c <- lapply(seeds, function(s) {
  p <- par_comp_model(num_comp = 1, seed = s)  # parameters for individua
  dat <- simulated_data(
    num_comp = 1,
    parameter = p,
    dose = 1200,
    seed = s
  )
  dat$seed <- s   # label/index for each dataset
  dat
})



datasets_2c <- lapply(seeds, function(s) {
  p <- par_comp_model(num_comp = 2, seed = s)
  dat <- simulated_data(
    num_comp = 2,
    parameter = p,
    dose = 1200,
    seed = s
  )
  dat$seed <- s
  dat
})


gp_data_list_1c <- lapply(datasets_1c, to_gp_data, dose = 1200)
gp_data_list_2c <- lapply(datasets_2c, to_gp_data, dose = 1200)


pk_data_list_1c <- lapply(datasets_1c, to_torsten_data)
pk_data_list_2c <- lapply(datasets_2c, to_torsten_data)
```




# Run and fit model

## PK model
```{r}
mod_pk <- cmdstan_model("two_compartment.stan")

init_2c <- function() list(
  CL = 10, Q = 15, V1 = 35, V2 = 105, ka = 2.5, sigma = 0.3
)

# Mis-specified case for pk2 model: Using pk2 model to fit pk1 data
fit_pk_1c <- lapply(pk_data_list_1c, function(pk_data) {
  mod_pk$sample(
    data = pk_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = init_2c,
    refresh = 50,
    show_messages = FALSE,
    show_exceptions = FALSE
  )
})

fit_pk_2c <- lapply(pk_data_list_1c, function(pk_data) {
  mod_pk$sample(
    data = pk_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = init_2c,
    refresh = 0,
    show_messages = FALSE,
    show_exceptions = FALSE
  )
})


```

### Inference (convergence, eﬀective sample size) for PK model
```{r}
pars_pk <- c("CL","Q","V1","V2","ka","sigma")

# Diagnostic for one-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_pk_1c), function(i) {
  s <- fit_pk_1c[[i]]$summary(variables = pars_pk)
  s$dataset_id <- i
  s
}))
subset(summ_table, rhat > 1.01 | ess_bulk < 200 | ess_tail < 200)

# Diagnostic for two-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_pk_2c), function(i) {
  s <- fit_pk_2c[[i]]$summary(variables = pars_pk)
  s$dataset_id <- i
  s
}))
subset(summ_table, rhat > 1.01 | ess_bulk < 200 | ess_tail < 200)
```

### Posterior predictive check for PK model
```{r}
compute_loo <- function(fit, log_lik_name = "log_lik") {
  # compute log_likihood draws
  ll_array <- fit$draws(variables = log_lik_name, format = "draws_array")
  ll_mat <- posterior::as_draws_matrix(ll_array)
  loo::loo(ll_mat)
}

loo_pk_1c <- lapply(fit_pk_1c, compute_loo, log_lik_name = "log_lik")
loo_pk_2c <- lapply(fit_pk_2c, compute_loo, log_lik_name = "log_lik")
```




## Gaussian Process
```{r}
mod_gp <- cmdstan_model("gp_model.stan")
make_init_gp <- function(gp_data) {
  function() list(
    a     = mean(log(gp_data$y_obs)),
    rho   = 1, #smoothness
    alpha = 1, #fluctuation 
    sigma = 0.3, #latent standard normal
    z     = rep(0, gp_data$N)
  )
}


fit_gp_1c <- lapply(gp_data_list_1c, function(gp_data) {
  mod_gp$sample(
    data = gp_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = make_init_gp(gp_data),
    refresh = 0,
    show_messages = FALSE,
  show_exceptions = FALSE
  )
})

fit_gp_2c <- lapply(gp_data_list_2c, function(gp_data) {
  mod_gp$sample(
    data = gp_data,
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    seed = 123,
    init = make_init_gp(gp_data),
    refresh = 0,
    show_messages = FALSE,
  show_exceptions = FALSE
  )
})


```



### Inference (convergence, eﬀective sample size)
```{r}
pars_gp <- c("a", "rho", "alpha", "sigma")

# Diagnostic for one-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_gp_1c), function(i) {
  s <- fit_gp_1c[[i]]$summary(variables = pars_gp)
  s$dataset_id <- i
  s
}))
# Filter: Rhat > 1.01
subset(summ_table, rhat > 1.01)
# Filter: ESS < 200
subset(summ_table, ess_bulk < 200 | ess_tail < 200)


# Diagnostic for two-compartment data
summ_table <- do.call(rbind, lapply(seq_along(fit_gp_2c), function(i) {
  s <- fit_gp_2c[[i]]$summary(variables = pars_gp)
  s$dataset_id <- i
  s
}))
# Filter: Rhat > 1.01
subset(summ_table, rhat > 1.01)
# Filter: ESS < 200
subset(summ_table, ess_bulk < 200 | ess_tail < 200)


```


### Posterior predictive check (LOO Diagnostics)

```{r}
loo_gp_1c <- lapply(fit_gp_1c, compute_loo, log_lik_name = "log_lik")
loo_gp_2c <- lapply(fit_gp_2c, compute_loo, log_lik_name = "log_lik")

```
There is an outlier in the simulated data has very large shape parameter $k$. Need to pinpoint the point and do exact LOO refit.

“最差的总是第 1 个数据点”其实非常合理：你 GP 用的观测数据里，第 1 个观测点几乎一定是 time = 0 那个 trough 点，而这个点在你现在的模拟机制下会天然变成 高杠杆 / 高敏感点，PSIS-LOO 就很容易把它标成最麻烦的点（Pareto k 最大）。
```{r}
loo::pareto_k_table(loo_gp_1c[[1]])
loo::pareto_k_table(loo_gp_2c[[1]])


k <- loo_gp_1c[[1]]$diagnostics$pareto_k
i_bad <- which(k > 1)
i_bad

```

```{r}
library(tidyverse)
library(patchwork)

# ==========================================
# 1. 批量评估 Model Fit (Posterior Median & 90% PI)
# ==========================================
plot_all_replicates <- function(fit_list_gp, fit_list_pk, data_list_gp, data_list_pk, num_reps = 5) {
  
  all_plot_data <- data.frame()
  
  for (i in 1:num_reps) {
    # ------ 提取 GP 模型数据 ------
    y_rep_gp <- fit_list_gp[[i]]$draws("y_rep", format = "matrix")
    df_gp <- data.frame(
      Replicate = paste("Replicate", i),
      Model = "GP",
      Time = data_list_gp[[i]]$t_obs,
      Obs = data_list_gp[[i]]$y_obs,
      Median = apply(y_rep_gp, 2, median),
      Lower = apply(y_rep_gp, 2, quantile, probs = 0.05),
      Upper = apply(y_rep_gp, 2, quantile, probs = 0.95)
    )
    
    # ------ 提取 PK 模型数据 ------
    y_rep_pk <- fit_list_pk[[i]]$draws("y_rep", format = "matrix")
    # 找到观测点对应的时间
    time_pk <- data_list_pk[[i]]$time[data_list_pk[[i]]$iObs]
    df_pk <- data.frame(
      Replicate = paste("Replicate", i),
      Model = "PK",
      Time = time_pk,
      Obs = data_list_pk[[i]]$cObs,
      Median = apply(y_rep_pk, 2, median),
      Lower = apply(y_rep_pk, 2, quantile, probs = 0.05),
      Upper = apply(y_rep_pk, 2, quantile, probs = 0.95)
    )
    
    # 合并当前 replicate 的数据
    all_plot_data <- bind_rows(all_plot_data, df_gp, df_pk)
  }
  
  # ------ 绘制一整张大图 ------
  p <- ggplot(all_plot_data, aes(x = Time)) +
    geom_ribbon(aes(ymin = Lower, ymax = Upper, fill = Model), alpha = 0.2) +
    geom_line(aes(y = Median, color = Model), linewidth = 0.8) +
    geom_point(aes(y = Obs), color = "black", size = 1) +
    facet_grid(Replicate ~ Model) + # 行是 Replicate, 列是 Model
    scale_fill_manual(values = c("GP" = "blue", "PK" = "red")) +
    scale_color_manual(values = c("GP" = "blue", "PK" = "red")) +
    labs(title = "Model Fit Comparison: GP vs PK (Median & 90% Predictive Interval)",
         y = "Concentration", x = "Time") +
    theme_bw() +
    theme(legend.position = "none", strip.text = element_text(size = 10, face = "bold"))
  
  return(p)
}

# 运行代码一次性画出 1-Compartment 数据的 5 个 Reps 比较图
final_plot_2c <- plot_all_replicates(fit_gp_2c, fit_pk_2c, gp_data_list_2c, pk_data_list_2c, num_replicates)
print(final_plot_2c)
```

```{r}
# ==========================================
# 2. 批量评估 Predictive Performance (ELPD Table)
# ==========================================
compare_elpd_all_reps <- function(loo_list_gp, loo_list_pk, num_reps = 5) {
  
  results_list <- lapply(1:num_reps, function(i) {
    l_gp <- loo_list_gp[[i]]
    l_pk <- loo_list_pk[[i]]
    
    # 提取 GP 的 ELPD 和 Standard Error (SE)
    elpd_gp <- l_gp$estimates["elpd_loo", "Estimate"]
    se_gp <- l_gp$estimates["elpd_loo", "SE"]
    
    # 提取 PK 的 ELPD 和 Standard Error (SE)
    elpd_pk <- l_pk$estimates["elpd_loo", "Estimate"]
    se_pk <- l_pk$estimates["elpd_loo", "SE"]
    
    # 计算 ELPD ± SD(SE) 区间是否重叠 (Proposal 中的要求)
    # 逻辑: 如果 A 的下界 <= B 的上界，且 B 的下界 <= A 的上界，则重叠
    gp_interval <- c(elpd_gp - se_gp, elpd_gp + se_gp)
    pk_interval <- c(elpd_pk - se_pk, elpd_pk + se_pk)
    intervals_overlap <- (gp_interval[1] <= pk_interval[2]) && (pk_interval[1] <= gp_interval[2])
    
    # 使用 loo_compare 进行精确比较
    comp <- loo::loo_compare(list(GP = l_gp, PK = l_pk))
    best_model <- rownames(comp)[1]
    
    # loo_compare 中，最好的模型在第一行且差值为0，第二行是两者的绝对差值
    elpd_diff <- abs(comp[2, "elpd_diff"])
    se_diff <- comp[2, "se_diff"]
    
    # 判断差异是否显著: 经验法则是 |ELPD_diff| > 2 * SE_diff
    is_significant <- elpd_diff > (2 * se_diff)
    
    # 整合当前 replicate 的统计结果
    data.frame(
      Replicate = i,
      Best_Model = best_model,
      ELPD_GP = round(elpd_gp, 2),
      SE_GP = round(se_gp, 2),
      ELPD_PK = round(elpd_pk, 2),
      SE_PK = round(se_pk, 2),
      Overlap_1SD = intervals_overlap, # 是否在彼此的 elpd +- sd 内
      ELPD_Diff = round(elpd_diff, 2),
      SE_Diff = round(se_diff, 2),
      Significant_Diff = is_significant
    )
  })
  
e
  final_table <- do.call(rbind, results_list)
  return(final_table)
}


elpd_comparison_1c <- compare_elpd_all_reps(loo_gp_1c, loo_pk_1c, num_reps = 5)


print(elpd_comparison_1c)
```

